Software architectures to integrate workflow engines in science gateways

### 2016-05-14 paper.tex
including section 3

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the potential of having
  more than one engine instance in the architecture simultaneously. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to means that adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines can be implemented through some kind
  of auto-scaling mechanism to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 means that
  adding a new engine instance requires some form of manual
  intervention, which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 means that new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  introduced by the architecture regarding task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not introduce any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are enabled by the architecture, and
  \texttt{O$_1$}=1 otherwise.
  \note{si}{why here we do not have the "possible to implement" situation?}

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow).  Fine-grained information about workflow tasks is
  required to properly troubleshoot workflow executions.
  \texttt{O$_2$}=0 means that such information is easy to access, and
  \texttt{O$_2$}=1 otherwise.
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is necessary
(\texttt{R$_2$}=1).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). 
Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen.

@@

\paragraph{Integration complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).

>>>>

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the potential of having
  more than one engine instance in the architecture simultaneously. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to means \note{vladimir}{is this the right word here, 'to means'?} that adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines can be implemented through some kind
  of auto-scaling mechanism to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 means that
  adding a new engine instance requires some form of manual
  intervention, which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 means that new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  introduced by the architecture regarding task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not introduce any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are enabled by the architecture, and
  \texttt{O$_1$}=1 otherwise.
  \note{si}{why here we do not have the "possible to implement" situation?}
  \note{vladimir}{I agree with Silvia, we should have the intermediate values as well, since meta-workflows might be straightforwardly supported by the gateway, or they are still possible to implement but only with some workarounds. Like it was with SHIWA: first 6 months coarsed-grained interoperability looked more like a workaround till it was better integrated and supported by WS-PGRADE itself}

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow).  Fine-grained information about workflow tasks is
  required to properly troubleshoot workflow executions.
  \texttt{O$_2$}=0 means that such information is easy to access, and
  \texttt{O$_2$}=1 otherwise. \note{vladimir}{otherwise - does it mean the fine-grained info is available but hard to access or not available it all? the latter happens as well}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is necessary
(\texttt{R$_2$}=1).
\note{vladimir}{there is a large piece of text in Scalability about reliability of the workflow engine task which on my opinion suits Robustness better}

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). 
Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure.\note{vladimir}{isn't this about robustness?} In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen.

@@

\paragraph{Integration complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\note{vladimir}{and if we have a meta-workflow with 3 and more sub-workflows the score will be even higher, is it a fair calculation for an architecture? we did not increase the scores for possible meta-workflows supported by wf engines in other cases}
